---
title: "Assignment 3: Data Exploration"
author: "Summer Heschong"
date: "Fall 2024"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on Data Exploration.

## Directions

1.  Rename this file `<FirstLast>_A03_DataExploration.Rmd` (replacing `<FirstLast>` with your first and last name).
2.  Change "Student Name" on line 3 (above) with your name.
3.  Work through the steps, **creating code and output** that fulfill each instruction. 
4.  Assign a useful **name to each code chunk** and include ample **comments** with your code.
5.  Be sure to **answer the questions** in this assignment document.
6.  When you have completed the assignment, **Knit** the text and code into a single PDF file.
7.  After Knitting, submit the completed exercise (PDF file) to the dropbox in Canvas.

**TIP**: If your code extends past the page when knit, tidy your code by manually inserting line breaks.

**TIP**: If your code fails to knit, check that no `install.packages()` or `View()` commands exist in your code. 

---

## Set up your R session

1.  Load necessary packages (tidyverse, lubridate, here), check your current working directory and upload two datasets: the ECOTOX neonicotinoid dataset (ECOTOX_Neonicotinoids_Insects_raw.csv) and the Niwot Ridge NEON dataset for litter and woody debris (NEON_NIWO_Litter_massdata_2018-08_raw.csv). Name these datasets "Neonics" and "Litter", respectively. Be sure to include the subcommand to read strings in as factors.

```{r,  message = FALSE}
#setting up
library(tidyverse); library(lubridate); library(here) #load packages

here() #check working directory

#load data sets:
Neonics <- read.csv(
  file = here("./Data/Raw/ECOTOX_Neonicotinoids_Insects_raw.csv"),
  stringsAsFactors = TRUE)   #load and name neonics

#view(Neonics) 

Litter <- read.csv(
  file = here("./Data/Raw/NEON_NIWO_Litter_massdata_2018-08_raw.csv"),
  stringsAsFactors = TRUE)   #load and name litter

#view(Litter) 
```

## Learn about your system

2.  The neonicotinoid dataset was collected from the Environmental Protection Agency's ECOTOX Knowledgebase, a database for ecotoxicology research. Neonicotinoids are a class of insecticides used widely in agriculture. The dataset that has been pulled includes all studies published on insects. Why might we be interested in the ecotoxicology of neonicotinoids on insects? Feel free to do a brief internet search if you feel you need more background information.

> Answer:We might be interested in this topic if we were researching certain insects or the effect of neonicotioids on ecosytems more generally. These chemicals are harmful to many beneficial insects like pollinators and aquatic invertebrates, both of which are integral to the health of their ecosytems. 


3.  The Niwot Ridge litter and woody debris dataset was collected from the National Ecological Observatory Network, which collectively includes 81 aquatic and terrestrial sites across 20 ecoclimatic domains. 32 of these sites sample forest litter and woody debris, and we will focus on the Niwot Ridge long-term ecological research (LTER) station in Colorado. Why might we be interested in studying litter and woody debris that falls to the ground in forests? Feel free to do a brief internet search if you feel you need more background information.

> Answer:So I actually was working at NEON before coming to school! I was on the flora team so I did both the litter and fine woody debris protocols. I'm so excited that we're using their data! Litter is very important in nutrient cylcling in a forest; so we might be interested to know how much carbon for example is being mineralized through the litter. 

4.  How is litter and woody debris sampled as part of the NEON network? Read the NEON_Litterfall_UserGuide.pdf document to learn more. List three pieces of salient information about the sampling methods here:

> Answer: 
 1.Elevated traps are set out that catch litter. The litter is collected from them in the spring and fall
 2.Ground traps are designated rectangles of ground where fine woody debris is collected once a year
 3. The samples of litter are taken back to the lab and sorted by type i.e. leaves, seeds, flowers, etc.

## Obtain basic summaries of your data (Neonics)

5.  What are the dimensions of the dataset?

```{r}
#dimensions of dataset?
colnames(Neonics) #column names of dataset
str(Neonics) #structure and dimensions of dataset

```

6.  Using the `summary` function on the "Effect" column, determine the most common effects that are studied. Why might these effects specifically be of interest? [Tip: The `sort()` command is useful for listing the values in order of magnitude...]

```{r}
#What are the most common effects that are studied?
sort(summary(Neonics$Effect)) 
```

> Answer:
Population is the most common effects studied. These are probably of specific interest because studying the population of an insect species can tell you a lot about the overall health and wellbeing of the species and its reaction to insecticide. 

7.  Using the `summary` function, determine the six most commonly studied species in the dataset (common name). What do these species have in common, and why might they be of interest over other insects? Feel free to do a brief internet search for more information if needed.[TIP: Explore the help on the `summary()` function, in particular the `maxsum` argument...]

```{r}
#6 most commonly studied species in dataset?
sort(summary(Neonics$Species.Common.Name))
help(summary) #what is a 'maxsum'?
sort(summary(Neonics$Species.Common.Name, maxsum = 670))
```

> Answer:The 6 most commonly studied species excluding the Other category  are the Honey Bee, the Parasitic Wasp, the Buff Tailed Bumblebee, The Carniolan Honey Bee, the Bumble Bee, and the Italian Honeybee. All of these except for the Parasitic Wasp are in the family Apidae. These might be of interest over other insects because the family Apidae contains many important pollinators, and you wouldn't want to use insecticide on your crops and kill the insects that pollinate them. Parasitic Wasps also prey on a lot of insects that are detrimental to agricultural crops so you would not want to harm them with insecticide either. 

8.  Concentrations are always a numeric value. What is the class of `Conc.1..Author.` column in the dataset, and why is it not numeric? [Tip: Viewing the dataframe may be helpful...]

```{r}
#Class of column?
str(Neonics$Conc.1..Author.)

```

> Answer:'Conc.1..Author.' is showing up as a factor. I think this is because the numbers in this column mean different things. Some of them are the concentrations of active ingredients and some are the concentration of the formula.

## Explore your data graphically (Neonics)

9.  Using `geom_freqpoly`, generate a plot of the number of studies conducted by publication year.

```{r}
#make plot of studies by pub. year
ggplot(Neonics) + geom_freqpoly(aes(x=Publication.Year))

```

10. Reproduce the same graph but now add a color aesthetic so that different Test.Location are displayed as different colors.

```{r}
#Represent test locations with different colors
ggplot(Neonics) + geom_freqpoly(aes(x=Publication.Year,
                                    color=Test.Location))
```

Interpret this graph. What are the most common test locations, and do they differ over time?

> Answer:The most common test locations are Lab and Field natural. They do differ over time. Lab is the most common between 2010-2020 and Field Natural is most common from around 2007-2010

11. Create a bar graph of Endpoint counts. What are the two most common end points, and how are they defined? Consult the ECOTOX_CodeAppendix for more information. 

[**TIP**: Add `theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))` to the end of your plot command to rotate and align the X-axis labels...]

```{r}
#bar graph of Endpoint counts
ggplot(Neonics, aes(x=Endpoint)) + geom_bar(); 
                theme(axis.text.x = element_text(,
                angle = 90, vjust = 0.5, hjust=1))

```

> Answer:The two most common endpoints are LOEL and NOEL. LOEL is the lowest concentration of insecticide that had an observable effect on the insects; while, NOEL is the highest concentration of insecticide that produced no observable effect on the insects. 

## Explore your data (Litter)

12. Determine the class of collectDate. Is it a date? If not, change to a date and confirm the new class of the variable. Using the `unique` function, determine which dates litter was sampled in August 2018.

```{r}
#What is the class of collectDate?
str(Litter$collectDate) #collectDate is a function
Litter$collectDate <- as.Date(Litter$collectDate, format = "%Y%m%d") #changed collectDate to a date
unique(Litter$collectDate)


```

13. Using the `unique` function, determine how many different plots were sampled at Niwot Ridge. How is the information obtained from `unique` different from that obtained from `summary`?

```{r}
#using the 'unique' function vs the 'summary' function
unique(Litter$plotID)
summary(Litter$plotID)
```

> Answer:12 plots were sampled at Niwot Ridge. The information obtained from 'unique' is the same data type as what you put into it but with the duplicate items gone. 'summary' shows you how many entries of each item there are in the data set. 

14. Create a bar graph of functionalGroup counts. This shows you what type of litter is collected at the Niwot Ridge sites. Notice that litter types are fairly equally distributed across the Niwot Ridge sites.

```{r}
#bar graph of funcionalGroup
ggplot(Litter, aes(x=functionalGroup)) + geom_bar()

```

15. Using `geom_boxplot` and `geom_violin`, create a boxplot and a violin plot of dryMass by functionalGroup.

```{r}
#boxplot and violin plot of dryMass by functionalGroup
ggplot(Litter, aes(y=dryMass, x=functionalGroup)) + geom_boxplot()
ggplot(Litter, aes(y=dryMass, x=functionalGroup)) + geom_violin()
```

Why is the boxplot a more effective visualization option than the violin plot in this case?

> Answer:This data is not the best for a violin plot because it shows up as lines. It looks much better in a box plot

What type(s) of litter tend to have the highest biomass at these sites?

> Answer:Needles tend to have the highest biomass
