---
title: "Assignment 5: Data Visualization"
author: "Summer Heschong"
date: "Fall 2024"
output: pdf_document
geometry: margin=2.54cm
fig_height: 3
fig_width: 4
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on Data Visualization 

## Directions
1. Rename this file `<FirstLast>_A05_DataVisualization.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure your code is tidy; use line breaks to ensure your code fits in the knitted output.
5. Be sure to **answer the questions** in this assignment document.
6. When you have completed the assignment, **Knit** the text and code into a single PDF file.

---

## Set up your session 

1. Set up your session. Load the tidyverse, lubridate, here & cowplot packages, and verify your home directory. Read in the NTL-LTER processed data files for nutrients and chemistry/physics for Peter and Paul Lakes (use the tidy `NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv` version in the Processed_KEY folder) and the processed data file for the Niwot Ridge litter dataset (use the `NEON_NIWO_Litter_mass_trap_Processed.csv` version, again from the Processed_KEY folder). 

2. Make sure R is reading dates as date format; if not change the format to date.

```{r initialize project}
#1 
library(tidyverse);library(lubridate);library(here); library(cowplot)
getwd()
here()
PeterPaul.chem.nutrients <- read.csv(here(
  'Data/Processed_KEY/NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv'),
  stringsAsFactors = TRUE)
Litter <- read.csv(here(
  'Data/Processed_KEY/NEON_NIWO_Litter_mass_trap_Processed.csv'),
  stringsAsFactors = TRUE)
#2 
PeterPaul.chem.nutrients$sampledate <- ymd(PeterPaul.chem.nutrients$sampledate)
Litter$collectDate <- ymd(Litter$collectDate)

```
## Define your theme
3. Build a theme and set it as your default theme. Customize the look of at least two of the following: 
 * Plot background
 * Plot title
 * Axis labels
 * Axis ticks/gridlines
 * Legend

```{r build plot theme}
#3
my_theme <-
  theme(
    panel.background = element_rect(fill = 'lightblue'),
    plot.title = element_text(
      hjust = 0.5, 
      size = 16, 
      face= 'bold', 
      colour = 'black')
  )

  
```

## Create graphs
For numbers 4-7, create ggplot graphs and adjust aesthetics to follow best practices for data visualization. Ensure your theme, color palettes, axes, and additional aesthetics are edited accordingly.

4. [NTL-LTER] Plot total phosphorus (`tp_ug`) by phosphate (`po4`), with separate aesthetics for Peter and Paul lakes. Add line(s) of best fit using the `lm` method. Adjust your axes to hide extreme values (hint: change the limits using `xlim()` and/or `ylim()`).

```{r plot total P vs PO4}
#4 
tpbypo4_plot <- PeterPaul.chem.nutrients %>%
  ggplot(aes(x=po4, y=tp_ug))+
  geom_point(aes(color=ifelse(lakename == 'Peter Lake', 'blue', 'red'))) +
  geom_smooth(method = 'lm',se=FALSE) +
  xlim(0, 20) +
  ylim(0, 100) +
  labs(color='Lake')  #I can't figure out how to change the labels 
#of the blue and red dots in the legend
tpbypo4_plot
```


5. [NTL-LTER] Make three separate boxplots of (a) temperature, (b) TP, and (c) TN, with month as the x axis and lake as a color aesthetic. Then, create a cowplot that combines the three graphs. Make sure that only one legend is present and that graph axes are aligned. 

Tips:
* Recall the discussion on factors in the lab section as it may be helpful here. 
* Setting an axis title in your theme to `element_blank()` removes the axis title (useful when multiple, aligned plots use the same axis values)
* Setting a legend's position to "none" will remove the legend from a plot.
* Individual plots can have different sizes when combined using `cowplot`.

```{r Create boxplots}
#5 
#convert 'month' to factor
PeterPaul.chem.nutrients$month <- factor(PeterPaul.chem.nutrients$month)

#Make boxplots for temp, tp, and tn, and remove x axis title and legend from all but one
Lake_boxplot_temp <- PeterPaul.chem.nutrients %>%
  ggplot(aes(x=month, y=temperature_C))+
  geom_boxplot(aes(color=lakename)) +
  theme(axis.title.x = element_blank())

Lake_boxplot_TP <- PeterPaul.chem.nutrients %>%
  ggplot(aes(x=month, y=tp_ug))+
  geom_boxplot(aes(color=lakename)) +
  theme(axis.title.x = element_blank(),
        legend.position = 'none')

Lake_boxplot_TN <- PeterPaul.chem.nutrients %>%
  ggplot(aes(x=month, y=tn_ug))+
  geom_boxplot(aes(color=lakename)) +
  theme(axis.title.x = element_blank(),
        legend.position = 'none')

#combine and align boxplots using cowplot
TpTnTemp_boxplot <- plot_grid(
  Lake_boxplot_temp, 
  Lake_boxplot_TN, 
  Lake_boxplot_TP,
  nrow=3, align='v', axis='x')
  
TpTnTemp_boxplot
```

Question: What do you observe about the variables of interest over seasons and between lakes? 

> Answer: The temperature is highest in late summer for both of the lakes. Tn and Tp are slightly higher in Peter Lake than Paul Lake. 

6. [Niwot Ridge] Plot a subset of the litter dataset by displaying only the "Needles" functional group. Plot the dry mass of needle litter by date and separate by NLCD class with a color aesthetic. (no need to adjust the name of each land use)

7. [Niwot Ridge] Now, plot the same plot but with NLCD classes separated into three facets rather than separated by color. 

```{r Plot litter}
#6
litter_plotcolor <- Litter %>%
  filter(functionalGroup=='Needles') %>%
  ggplot(aes(x=collectDate, y=dryMass, color=nlcdClass)) +
  geom_point()
litter_plotcolor
#7
litter_plotfacet <- Litter %>%
  filter(functionalGroup=='Needles') %>%
  ggplot(aes(x=collectDate, y=dryMass)) +
  geom_point() +
  facet_wrap(vars(nlcdClass), nrow=3)
litter_plotfacet

```
Question: Which of these plots (6 vs. 7) do you think is more effective, and why? 

> Answer:I think that plot 7 is more effective because in plot 6 the points overlap and hide each other so it's hard to see all the data. 